\chapter{Architecture}
\label{chap:architecture}

\section{Invariants}

The project is guided by a small set of invariants:

\begin{enumerate}
    \item Each piece of application service state is owned by one shard.
    \item Only the owning shard may mutate its service state.
    \item Application services should not be modeled as shared
          \texttt{Arc<Mutex<Service\textgreater{}\textgreater{}} state.
    \item Cross-shard interaction happens through typed messages, typed reply
          handles, or typed submitter calls.
    \item Values crossing shard boundaries are owned.
    \item References to shard-local state must not escape synchronous access
          closures and must not cross \texttt{.await}.
    \item Async work remains on its assigned shard unless explicitly submitted
          elsewhere.
    \item Observability snapshots are owned values, not borrowed runtime
          internals.
    \item Unsafe code is isolated behind small safe APIs.
\end{enumerate}

\section{The shared-nothing rule}

The central design rule: application state has one owner. In the
standard-library services that owner is a shard thread receiving typed
commands. In the async runtime it is an executor shard polling futures.
The form changes; the rule does not.

This is why the project avoids \texttt{Arc<Mutex<Service\textgreater{}\textgreater{}}.
A mutex makes mutation mechanically safe but hides the question the project
studies: which shard owns this state, where does work run, and when does the
ownership boundary get crossed?

Shared-nothing does not mean synchronization-free. Wakers, reply state,
lifecycle flags, counters, and submission queues use synchronization.
Application service state does not. It stays local to its shard.

\begin{notabene}[Why avoid a service mutex?]
A service mutex answers only one question: can two threads mutate this value at
the same time? The runtime is asking a stronger question: which shard is
responsible for this value, which work is allowed to touch it, and how does
another shard request a change? Keeping service state shard-owned preserves
that answer in the program structure.
\end{notabene}

\begin{center}
\begin{tikzpicture}[
    box/.style={draw=sitasMuted, fill=sitasCodeBackground,
        minimum width=48mm, minimum height=8mm, font=\footnotesize\ttfamily,
        rounded corners=1.5pt},
    arr/.style={->, >=stealth, thick, color=sitasAccent},
    lbl/.style={font=\tiny\itshape, color=sitasMuted}
]
\node[box] (caller) at (0, 0) {Caller thread or task};
\node[box] (shard)  at (0, -1.8) {Owning shard};
\node[box] (reply)  at (0, -3.6) {Owned reply or snapshot};

\draw[arr] (caller) -- node[right, lbl, align=left] {owned command /\\owned future /\\typed submitter call} (shard);
\draw[arr] (shard) -- node[right, lbl] {synchronous local mutation} (reply);
\end{tikzpicture}
\end{center}

\section{Boundaries and lifetimes}

Most runtime bugs are boundary bugs. A value crosses to another thread without being owned. A reference to local state escapes a
closure. A future keeps a borrowed value across \texttt{.await}. A completion
operation drops a buffer before the kernel has finished with it.

\texttt{sitas} tries to make those boundaries visible:

\begin{description}
    \item[Cross-shard transfer] uses owned values and explicit
        \texttt{Send + 'static} requirements where a future can move to another
        thread.
    \item[Shard-local access] gives a closure synchronous access to
        \texttt{\&mut T}; the closure returns an owned result and cannot
        suspend.
    \item[Observability] returns owned snapshots. A monitoring thread can keep
        a snapshot after the runtime has moved on.
    \item[Kernel I/O] stores owned buffers until completions are observed or
        until abandoned operation cleanup proves they can be discarded.
\end{description}

\section{Why owned snapshots}

Observability could expose references into executor internals, but that would
make a diagnostic API participate in the same lifetime and locking problems as
the runtime. The project treats observation as a copy boundary. A snapshot is a
point-in-time value that can be sorted, printed, diffed, or sent to another
thread without keeping the executor borrowed.

This matters because long-running work needs inspection while it is still
changing. Owned snapshots let monitoring code be ordinary application code.
They also prevent a debugging tool from accidentally extending the lifetime of
task, timer, or shard-local internals.

\section{Layered model}

The current runtime can be read as a layered system:

\begin{lstlisting}[language={},caption={Conceptual layers}]
Application service APIs
    |
Typed command/reply and submitter APIs
    |
Placement and shard routing
    |
Shard-local service state / ShardLocal<T>
    |
Shard executor or std shard thread
    |
Mailbox, task queue, timers, readiness, wakeups
    |
OS primitives: pipe/poll/epoll/io_uring/affinity
\end{lstlisting}

The standard-library baseline exercises the upper part of the model. The
custom executor and OS runtime layers fill in the lower layers while preserving
the same shared-nothing service direction.

\section{Primary modules}

\begin{description}
    \item[\texttt{runtime}] The reusable standard-library kernel for bounded
        shard mailboxes, reply handles, shard startup, and shutdown.
    \item[\texttt{kv}] The reference sharded key-value service.
    \item[\texttt{counter}] A smaller second service proving that the runtime
        primitives are not key-value-specific.
    \item[\texttt{sharded}] A generic \texttt{ShardService} trait and
        \texttt{Sharded<S>} runtime providing common lifecycle (start, shutdown,
        snapshot) for user-defined shard-local services. Shutdown is automatic
        via \texttt{stop\_command}.
    \item[\texttt{placement}] Key-to-shard routing.
    \item[\texttt{stream\_reply}] Streaming reply channels. A
        \texttt{StreamProducer<T>} delivers a sequence of owned values. Two
        async patterns: \texttt{StreamBatch<'a, T>} borrows the reply for
        \texttt{next\_batch()} loops, \texttt{StreamFuture<T>} owns the reply
        with \texttt{collect()} and \texttt{fold()} convenience methods.
        Senders use an atomic refcount so intermediate clone drops do not
        prematurely close the stream.
    \item[\texttt{async\_service}] Async wrappers bridging std-layer services
        (\texttt{ShardedKv}, \texttt{ShardedCounter}) into the executor via
        \texttt{submit\_* + reply.wait\_async().await}.
    \item[\texttt{os}] Unix FFI, wake pipes, readiness backends, affinity, and
        Linux \texttt{io\_uring} experiments. Readiness backends now cover
        \texttt{epoll} (Linux), \texttt{kqueue} (macOS/iOS/FreeBSD/NetBSD/
        OpenBSD), and \texttt{poll} (other Unix).
    \item[\texttt{executor}] The single-threaded async kernel. Sub-modules
        include \texttt{tcp}, \texttt{udp}, \texttt{backpressure}, and
        Linux \texttt{io\_uring} dispatch. \texttt{Spawner::with\_backpressure}
        integrates a counting semaphore into the spawn path.
    \item[\texttt{sharded\_executor}] One executor per shard thread, plus
        submission and observation APIs.
    \item[\texttt{shard\_local}] One owned value per executor shard with
        checked local access. Provides \texttt{with\_current},
        \texttt{with\_current\_result}, \texttt{get\_current},
        \texttt{map\_all}, and worker/scope APIs.
    \item[\texttt{shard\_mailbox}] Typed owned-message transfer between
        shards. Provides uniform key-to-shard routing through
        \texttt{ShardMailboxSet<M>} and non-uniform logical work-unit routing
        through \texttt{WorkUnitMailboxSet<N, M>}.
    \item[\texttt{sharded\_tcp}] Network-facing sharded TCP server with
        \texttt{SO\_REUSEPORT} on Linux and single-accept distribution on
        other platforms.
    \item[\texttt{metrics}] Thread-safe atomic-counter metrics collector with
        owned \texttt{MetricsSnapshot} for observability.
\end{description}

\section{How to trace a feature through the layers}

When learning or extending the runtime, trace one feature down the stack. For
example, a named task in a scheduling group passes through several layers:

\begin{enumerate}
    \item user code calls a typed spawn API;
    \item the spawner validates that the scheduling group belongs to this
          executor;
    \item a task is allocated with a name and group id;
    \item the scheduler places the task into the group's ready queue;
    \item the executor polls the selected task;
    \item the task records status, wait interest, poll count, and timestamps;
    \item a snapshot later resolves the group id back to an owned group name.
\end{enumerate}

This kind of trace is more useful than reading modules in isolation. It shows
which layer owns validation, scheduling, and observability, and where the
public API remains safe.
