\chapter{Executor}

\section{Role}

The executor is a minimal single-threaded async kernel. It owns pinned futures
inside tasks, maintains a ready queue, coalesces repeated wakes, and drives a
root future through \texttt{run\_until} while also polling spawned tasks.

The executor deliberately avoids Tokio or another third-party runtime. That
keeps the interesting internals visible: wakers, task queues, timers, readiness
registration, cancellation, and completion dispatch all live in this code base.

\section{Why a custom executor}

The custom executor exists because the project needs to control the meaning of
``where work runs.'' In a shard-per-core runtime, a future is not just an async
computation; it is also part of a placement decision. The executor must be able
to say that this task belongs to this shard, this scheduling group, this ready
queue, and this local reactor state.

A third-party runtime would solve many engineering problems, but it would also
move the most interesting questions behind someone else's scheduler. Here the
small executor is the laboratory. It makes it possible to inspect how a waker
requeues one task, how a dropped timeout removes timer state, how a TCP future
records fd interest, and how a completion future keeps buffers alive.

\section{Responsibilities}

\begin{itemize}
    \item Poll ready tasks with a budget so self-waking tasks cannot starve
          timers and I/O forever.
    \item Implement custom task wakers that enqueue tasks onto the local ready
          queue.
    \item Catch task panics at executor boundaries.
    \item Preserve panic payloads where join handles and \texttt{block\_on}
          make them observable.
    \item Provide \texttt{sleep}, \texttt{timeout}, \texttt{race}, and
          \texttt{yield\_now}.
    \item Provide typed join handles and abort support.
    \item Provide cooperative stop tokens, \texttt{Notify}, and
          \texttt{TaskScope}.
    \item Drive readiness futures for read and write interests.
    \item On Linux, drive executor-owned \texttt{io\_uring} read/write-at
          futures.
\end{itemize}

\section{Task lifecycle}

A spawned future is stored inside a task. The task has a stable executor-local
\texttt{TaskId}, an optional name, a scheduling group id, and internal state
describing whether it is queued, polling, waiting, completed, or cancelled.

The lifecycle is intentionally explicit:

\begin{lstlisting}[language={},caption={Simplified task lifecycle}]
spawn
  -> create Task with pinned Future
  -> queue Task in a scheduling group
  -> poll Future
       Ready(())  -> completed
       Pending    -> waiting until waker fires
  -> wake
  -> queue Task again
\end{lstlisting}

A repeated wake does not create repeated queue entries for the same task. The
task records that it is already queued, so repeated wakeups coalesce. This is a
small but important executor property: a chatty future should not multiply
itself in the ready queue.

Cancellation is normal future dropping. Aborting a join handle or dropping an
executor eventually drops the owned future. Timer and readiness futures must
therefore clean up their registrations in \texttt{Drop}; otherwise the reactor
would keep stale interests for tasks that no longer exist.

\begin{notabene}[Why cancellation is treated as ordinary]
Rust futures are cancelled by being dropped. The runtime cannot rely on a
separate cancellation callback running later. Cleanup must therefore be owned by
the future, task, join handle, or dispatcher state that is being dropped. This
is why many tests exercise timeout, abort, and dropped-future paths rather than
only successful completion.
\end{notabene}

\section{Wakers and the ready queue}

The custom waker does one job: make the task runnable again on the same
executor. It does not move the task to another shard, does not poll the task
directly, and does not borrow application state. It marks the task queued and
hands it back to the scheduler.

The executor loop then decides when to poll ready work. It does so in batches
bounded by \texttt{READY\_POLL\_BUDGET}. That budget is a fairness valve. A
self-waking task can keep itself ready, but it cannot prevent the executor from
checking timers, readiness, and completions forever.

The ready queue is intentionally executor-local. Cross-thread wakeups may need
to notify an executor that work became ready, but the task itself is still
polled by the executor that owns it. This keeps local task state, scheduling
group accounting, and reactor registrations aligned.

\section{Scheduling groups}

Scheduling groups are the first resource-class mechanism in the executor. Each
group owns a ready queue and has a relative share count. Ordinary spawn calls
use the default group. Explicit group APIs let code classify work such as
foreground request handling and background maintenance.

The reason to add groups before more advanced load balancing is that resource
classification is a local scheduling problem. A shard can decide how to divide
its own polling time among foreground and background tasks without moving
application state to another shard. That makes scheduling groups a natural fit
for the shared-nothing model.

The selection rule is deliberately simple. The scheduler chooses a non-empty
group with the lowest weighted virtual runtime, pops one task from that
group's FIFO queue, polls it, and charges the group for the observed wall-clock
poll duration. The charge is scaled by the group's shares.

\begin{lstlisting}[language={},caption={Weighted group selection}]
for each poll:
    group = non_empty_group_with_lowest_virtual_runtime()
    task = group.pop_front()
    duration = poll(task)
    group.virtual_runtime += duration * default_shares / group.shares
\end{lstlisting}

This is not preemption. A long poll still blocks the executor thread until the
future returns \texttt{Poll::Pending} or \texttt{Poll::Ready}. The mechanism is
cooperative: futures must yield, await I/O, await timers, or otherwise return
control to the executor.

Scheduling group handles are executor-local. A group created by one executor is
rejected by another. The default group handle is special because it names the
built-in default queue in any executor. The same idea appears in the sharded
runtime as a \texttt{ShardedSchedulingGroup}, which is one executor-local group
per shard plus a runtime identity check.

\section{Structured child tasks}

\texttt{TaskScope} groups child tasks under one owner. Dropping the scope
requests cooperative stop and aborts still-owned children. Calling
\texttt{shutdown} asks children to stop and waits for them. Calling
\texttt{shutdown\_timeout} bounds that wait and aborts uncooperative children.

Scopes can also spawn children into scheduling groups. This matters when a
server accepts work that should be both lifetime-scoped and resource-classified.
The scope does not implement scheduling itself; it delegates group validation
and task creation to the underlying \texttt{Spawner}.

\section{Wait ordering}

Each loop iteration polls ready work, wakes expired timers, dispatches locally
available \texttt{io\_uring} completions, and then decides how to wait.

Pending \texttt{io\_uring} work now takes the executor wait slot when the shard
has no ready tasks. That prevents unrelated timers or readiness interests from
delaying completion-based file I/O. This is still not the final form: the long
term goal is a single deadline-aware wait path that can combine timers,
readiness, and completion waiting without giving any source accidental priority.

\section{Observability}

Executor snapshots are owned values. They include task counts, named task
state, polling counters, timer counts, readiness counts, and Linux
\texttt{io\_uring} dispatcher counters when the dispatcher is installed.

Task snapshots include the task name, scheduling group id and name, lifecycle
state, wait interest, poll count, accumulated poll time, and key timestamps.
Helper methods derive task age and current state duration from a caller-supplied
\texttt{Instant}. This keeps the snapshot itself factual while making progress
views easy to write.

\begin{lstlisting}[language={},caption={Reading a task snapshot}]
task name:              useful for humans
status:                 queued, polling, waiting, completed, cancelled
waiting_for:            timer, readable fd, writable fd, unknown, none
poll_count:             how many times the future was polled
total_poll_time:        how much executor time the task consumed
state_duration_at(now): how long it has been in the current state
\end{lstlisting}
