\chapter{Introduction}
\label{chap:introduction}

\section{Purpose}

\texttt{sitas} is an experimental Rust runtime and service model inspired by
Seastar's shard-per-core architecture. The project is about understanding the 
runtime internals: how work is owned by shards, how cross-shard messages 
move, how a small executor drives futures, and how OS completion and 
readiness mechanisms can be exposed without depending on a third-party 
async runtime.

The code base started with a standard-library-only sharded key-value store. The
current branch extends that baseline with a custom executor, Unix readiness
primitives, TCP helpers, sharded async execution, shard-local values,
observability snapshots, CPU placement, and experimental Linux \texttt{io\_uring}
support.

\section{Prerequisites}

This manual assumes familiarity with:

\begin{itemize}
    \item Rust ownership, borrowing, and lifetimes;
    \item \texttt{Future}, \texttt{Poll}, \texttt{Waker}, and \texttt{Pin};
    \item basic concurrency: threads, channels, mutexes;
    \item Unix file descriptors and non-blocking I/O (for the I/O chapters).
\end{itemize}

\section{How to read this project}

Application code sees typed, owned boundaries. A caller asks a shard to do 
work. The owning shard mutates its local state and returns an owned reply. The 
implementation avoids hiding service state behind a global mutex.

The runtime keeps that model honest:

\begin{enumerate}
    \item futures are owned by executor tasks;
    \item tasks are assigned to one executor shard;
    \item wakers only requeue tasks on that executor;
    \item cross-shard work must be submitted explicitly;
    \item shard-local values can be borrowed only inside synchronous closures;
    \item observability returns copies, not borrowed internals.
\end{enumerate}

This manual follows that path. \autoref{chap:architecture} explains the
ownership rules. \autoref{chap:executor} explains how futures become runnable
work. \autoref{chap:sharding} explains how one executor becomes many
independent shards. \autoref{chap:io-runtime} explains why readiness and
completion are different mechanisms and how both appear in the runtime.

\begin{notabene}[Learning goal]
If you can explain where a future is polled, who owns the state it mutates,
what wakes it, and what happens when it is dropped, you understand the most
important parts of this code base.
\end{notabene}

\section{Why the project has this shape}

The runtime is built from small visible pieces. A mature runtime hides task
allocation, reactor registration, cross-thread wakeups, cancellation, and
kernel ownership behind polished APIs. That works in production but obscures
the machinery.

\texttt{sitas} therefore accepts some explicitness that a general-purpose
runtime might smooth over:

\begin{description}
    \item[Explicit shards] make ownership visible. If work runs on shard 2,
        shard 2 owns the mutation.
    \item[Explicit submitters] make cross-shard movement visible. A future does
        not silently migrate just because another shard is less busy.
    \item[Explicit snapshots] make observation visible. Monitoring is a read of
        owned runtime facts, not a borrowed view into live internals.
    \item[Explicit OS boundaries] make platform behavior visible. A readiness
        wait, an \texttt{io\_uring} completion, and a CPU affinity request are
        different contracts with the operating system.
\end{description}

This is the main design tradeoff. The project prefers APIs that teach the 
boundary they cross over APIs that hide it.

\section{Scope of this manual}

This manual is a LaTeX companion to the Markdown architecture notes. It is
intended for longer-form system documentation and for keeping a printable view
of the evolving runtime design.

It focuses on:

\begin{itemize}
    \item architectural invariants;
    \item executor and scheduling behavior;
    \item shard ownership and cross-shard submission;
    \item readiness, \texttt{epoll}, and \texttt{io\_uring} integration;
    \item example programs that exercise the runtime;
    \item build and validation operations.
\end{itemize}

\begin{notabene}
The project is still deliberately experimental. APIs are allowed to move when
that exposes cleaner runtime internals or a better teaching experience.
\end{notabene}

\section{What this project is not}

\texttt{sitas} is not trying to become a production replacement for mature
Rust runtimes. It is also not trying to copy Seastar. The point is to keep 
the mechanisms small enough that they can be inspected:

\begin{itemize}
    \item no Tokio, async-std, Glommio, Monoio, or actor framework is used as
          the core runtime;
    \item blocking standard-library service APIs remain distinct from
          awaitable executor APIs;
    \item unsafe and FFI code is kept near the operating-system boundary;
    \item mechanisms are introduced only after their semantics can be tested.
\end{itemize}

This makes the project useful as a learning resource. It has enough machinery
to demonstrate timers, wakers, readiness, \texttt{io\_uring}, shard-local
values, scheduling groups, and CPU placement while keeping those mechanisms
visible in ordinary Rust modules.

\section{Further reading}

\begin{description}
    \item[Seastar] The C++ shard-per-core framework that inspired the
        ownership model.\\
        \url{https://seastar.io}
    \item[ScyllaDB shard-per-core architecture] Practical discussion of
        shared-nothing database design on top of Seastar.\\
        \url{https://www.scylladb.com/product/technology/shard-per-core-architecture/}
    \item[Glommio] A Rust thread-per-core runtime using \texttt{io\_uring},
        useful for comparing design choices.\\
        \url{https://github.com/DataDog/glommio}
\end{description}
