\chapter{I/O Runtime}
\label{chap:io-runtime}

\section{Readiness}

The readiness path is small. A future registers interest in a file descriptor,
the scheduler records the waker, the Unix reactor sleeps using Linux
\texttt{epoll}, macOS/iOS \texttt{kqueue}, or fallback \texttt{poll} plus a
wake pipe, and readiness wakes the interested task.

This path is readiness-based, not completion-based. It is used by the current
TCP helpers and remains separate from the experimental \texttt{io\_uring}
completion path.

Readiness is the first I/O layer because it matches Unix non-blocking sockets
and is portable. Ownership stays simple: user code owns the stream or listener;
the runtime only tracks which task to wake when the descriptor may progress.

Readiness means the OS reports an operation is likely to make progress, not
that bytes are available. A readable stream may return fewer bytes than
requested. A writable stream may accept only part of a buffer. The future must
attempt the operation, handle \texttt{WouldBlock}, and re-register interest.

\begin{lstlisting}[language={},caption={Readiness future shape}]
poll future:
    try non-blocking operation
    if operation succeeds:
        return Ready(result)
    if operation would block:
        register fd interest and waker
        return Pending

reactor wait:
    OS reports fd readiness
    scheduler wakes interested task
\end{lstlisting}

Readiness futures do not own file descriptors. The caller owns the stream or
listener; the future borrows it for the async operation and registers interest
while pending.

A connection handler owns a stream and calls helpers against it in sequence.
The reactor is the waiting mechanism, not the connection owner.

\section{TCP helpers}

The TCP helpers are built on top of readiness futures rather than hiding a
separate runtime. They demonstrate the normal async server shape:

\begin{enumerate}
    \item put the listener or stream in non-blocking mode;
    \item accept or connect using readiness futures;
    \item spawn one handler task per accepted connection;
    \item use \texttt{TaskScope} or join handles to wait for handlers;
    \item propagate handler errors as ordinary \texttt{io::Result} values.
\end{enumerate}

Grouped TCP helper variants classify accepted handler tasks into scheduling
groups. The accept loop remains in the caller's task; only per-connection
handler work enters the requested group. This separates foreground request
handling from background maintenance or low-priority connections.

Accepting connections and serving connections have different lifecycle and
scheduling needs. The accept loop is one long-lived task. Handlers are many
shorter-lived tasks. Keeping those roles separate makes shutdown,
observability, and scheduling group assignment straightforward.

\section{Sharded TCP socket options}

\texttt{ShardedTcpServer} keeps Linux socket placement explicit. On Linux,
\texttt{SO\_REUSEPORT} lets each shard bind its own listener so the kernel can
distribute incoming connections across shard-local accept loops. The server can
also opt into \texttt{SO\_INCOMING\_CPU} with
\texttt{ShardedTcpIncomingCpu}. This option is disabled by default; callers
must request either sequential placement over \texttt{available\_cpu\_ids()} or
an explicit CPU list.

\texttt{SO\_INCOMING\_CPU} is socket metadata, not a polling mode. It does not
consume a kernel thread. It should not be confused with mechanisms such as
\texttt{io\_uring} \texttt{SQPOLL}, where a kernel polling thread may be part
of the setup. In this runtime, \texttt{SO\_INCOMING\_CPU} is only an optional
listener \texttt{setsockopt}; the accept, read, and write path remains
readiness-driven.

Accepted sharded TCP connections expose a Linux kTLS ULP helper. That helper
attaches the kernel \texttt{tls} upper-layer protocol to the socket. It is not
a complete TLS implementation: the handler or a TLS stack must still complete
the handshake and install the required kernel TX/RX crypto state. The server
does not own certificates, handshake transcripts, traffic secrets, or TLS
record policy, so it deliberately leaves the full kTLS handoff outside the
generic accept loop.

\begin{notabene}[Readiness lesson]
Readiness is a notification that retrying may be useful. Completion is a
notification that the operation has already finished. The code keeps those
concepts separate because the ownership and cancellation rules are different.
\end{notabene}

\section{\texttt{io\_uring}}

The Linux \texttt{io\_uring} backend is experimental. It currently supports a
narrow executor integration for owned-buffer \texttt{read\_at},
\texttt{read\_exact\_at}, and \texttt{write\_at} style file I/O.

\texttt{io\_uring} is included because it exercises a harder runtime problem
than readiness. The kernel can own an operation after the Rust future that
submitted it has been dropped. That forces the runtime to model operation ids,
owned buffers, late completions, and abandonment---exactly the internals this
project wants to make visible.

Host policy is part of the model. Many Linux deployments, especially
containers and hardened hosts, restrict or disable \texttt{io\_uring} because
it exposes a broad kernel/user boundary. Newer kernels are adding more
fine-grained restriction mechanisms, including cBPF-style filtering of allowed
\texttt{io\_uring} operations, so a host may allow a small opcode set rather
than enabling every operation. This runtime does not install or manage such
policy. It keeps the operations narrow and reports setup or submission failure
as unavailable or as an ordinary I/O error, depending on where the host
rejects the request.

Each Linux executor run loop installs a thread-local dispatcher when the host
allows ring setup. Executor-backed \texttt{io\_uring} futures store operation
ids and owned buffers. When polled on a shard, they queue an operation against
the thread-local dispatcher and register their task waker.

When the executor becomes idle, it submits pending ring entries and polls the
ring descriptor through the same reactor wait used for fd readiness and timer
deadlines. Ring descriptor readiness is only a notification that completion
queue entries may be harvested. The completed operation still flows through the
dispatcher, which matches operation ids, returns owned buffers, and wakes
waiting futures.

The dispatcher tracks:

\begin{itemize}
    \item queued submissions;
    \item locally buffered completions;
    \item registered wakers;
    \item abandoned operations;
    \item deferred owned buffers;
    \item cumulative dispatched, buffered, woken, and discarded operation
          counters.
\end{itemize}

\section{Completion lifecycle}

The \texttt{io\_uring} path is completion-based. A future submits an operation
and later receives a completion. That changes the ownership problem. With
readiness, the future borrows a stream and retries ordinary system calls. With
\texttt{io\_uring}, the kernel may hold a pointer to an owned buffer until the
completion arrives.

The runtime tracks operation state explicitly:

\begin{lstlisting}[language={},caption={Simplified completion lifecycle}]
future first poll
    allocate operation id
    move owned buffer into dispatcher tracking
    submit operation to ring
    register task waker
    return Pending

completion arrives
    record result and completed buffer
    wake registered task

future next poll
    take completion
    return Ready(result, buffer)
\end{lstlisting}

If the future is dropped before completion, the operation may still be in the
kernel. The dispatcher marks it abandoned and defers buffer cleanup until a
completion or safe discard point. This is one of the most important differences
between readiness and completion: cancellation involves kernel-owned in-flight
work, not just waker removal.

The dispatcher is a lifecycle object, not just a syscall wrapper. A thin
wrapper around \texttt{io\_uring\_enter} would not suffice for async Rust
integration. The runtime needs somewhere to hold buffers, match completions to
futures, wake tasks, and account for futures that vanished before the kernel
answered.

\section{Current limitation}

Timers, readiness waits, and \texttt{io\_uring} completions now share one
executor idle wait shape on Linux. This is still a staged integration rather
than a production event loop. Completion dispatch has a fixed internal
per-tick budget that is exposed in executor snapshots. Those snapshots also
count non-empty completion dispatch batches, dispatched completions, and
completion budget exhaustion events. They also expose an executor-owned
\texttt{io\_uring} lifecycle status, so callers can distinguish not-started,
unavailable, installed, and shutdown states. Strict availability requests via
\texttt{SITAS\_REQUIRE\_IO\_URING=1} fail executor setup instead of silently
falling back to unavailable. Executor teardown makes a bounded attempt to drain
abandoned completion work before discarding the thread-local dispatcher when no
live task wakers remain registered. The final dispatcher snapshot records
whether that drain completed, timed out, or was skipped because live wakers
remained. When a root \texttt{run\_until} future completes with spawned
\texttt{io\_uring} work still pending, the dispatcher remains installed for
that executor on the current thread so later executor calls can continue the
same operations. The runtime still does not expose a portable completion
source, tunable completion policy, or a deeper shutdown policy for mixed
readiness and completion workloads.

\section{Running statistics}

\href{https://en.wikipedia.org/wiki/Algorithms_for_calculating_variance\#Welford's_online_algorithm}{Welford's
algorithm} computes mean, variance, and standard deviation in a single pass
without storing individual samples. The \texttt{running\_stats} module provides
\texttt{RunningStatistics} as a small online accumulator useful for runtime
diagnostics and benchmark comparisons.

\begin{lstlisting}[language={},caption={RunningStatistics use}]
let mut stats = RunningStatistics::new();
stats.add_sample_duration(elapsed);       // record as seconds (f64)
stats.add_sample_interval(start, end);    // convenience for Instant pairs
stats.merge(&other);                      // Chan et al. parallel merge

// Accessors return Option<f64> (None until enough samples):
stats.count()         // u64
stats.min() / max()   // Option<f64>
stats.mean()          // Option<f64>
stats.variance()      // Option<f64>,  sample variance (n-1), n >= 2
stats.std_dev()       // Option<f64>
stats.cv()            // Option<f64>,  coefficient of variation in %
stats.total()         // u64 integral total
\end{lstlisting}

Duration convenience methods (\texttt{mean\_duration()},
\texttt{min\_duration()}, \texttt{max\_duration()},
\texttt{std\_dev\_duration()}) return \texttt{Option<Duration>}
instead of raw seconds.

The \texttt{merge} method uses Chan et al.'s parallel algorithm so partial
accumulators from different threads or shards can be combined without
reprocessing the original samples. This is useful when each shard collects
statistics independently and the caller wants aggregate numbers.

The display format prints count, mean, standard deviation, CV, min, max, and
total in a single human-readable line, which is how the index build examples
report per-shard timing statistics.

CV is unit-free. It expresses dispersion relative to the mean, making it useful
for comparing the stability of metrics with different scales (e.g.\ partition
scan times versus merge round times).
