\chapter{Operations}

\section{Building this manual}

The manual lives in \filepath{docs/system}. Build it with:

\begin{lstlisting}[language=bash]
cd docs/system
make
\end{lstlisting}

The output is \filepath{docs/system/sysdoc.pdf}. The build uses
\texttt{pdflatex} twice so the table of contents and references settle.

If the build fails, first check that the LaTeX distribution has the standard
packages used by \filepath{sitasdoc.cls}: \texttt{geometry}, \texttt{fancyhdr},
\texttt{titlesec}, \texttt{titletoc}, \texttt{listings}, \texttt{xcolor}, and
\texttt{hyperref}. The manual intentionally avoids project-specific generated
assets so it can be rebuilt from text alone.

Keeping the manual build simple is intentional. The documentation should not
depend on the runtime being runnable, Docker being available, or generated
diagrams being checked out from another tool. A plain text-to-PDF path makes it
cheap to update design explanations in the same commits as code changes.

\section{Cleaning generated files}

\begin{lstlisting}[language=bash]
cd docs/system
make clean
make distclean
\end{lstlisting}

\texttt{make clean} removes LaTeX auxiliary files. \texttt{make distclean} also
removes the generated PDF.

\section{Linux validation}

Some runtime behavior is Linux-specific. The Docker helper mounts the project
directory into a Linux container and can request an \texttt{io\_uring}-capable
configuration:

\begin{lstlisting}[language=bash]
SITAS_DOCKER_IO_URING=1 tools/linux-docker.sh cargo test os::uring::tests -- --nocapture
\end{lstlisting}

When the host or container configuration cannot provide \texttt{io\_uring}, the
Linux-only examples report that the backend is unavailable rather than faking a
successful run.

The Docker path exists because macOS is a good development environment for most
of the code, while Linux is the only honest place to test Linux primitives such
as hard CPU affinity and \texttt{io\_uring}. The helper keeps that platform
check repeatable without pretending the project is platform-neutral in areas
where it is not.

\section{Continuous integration}

The project includes a GitHub Actions workflow (\filepath{.github/workflows/ci.yml})
that runs on every push and pull request:

\begin{itemize}
    \item \texttt{rustfmt} and \texttt{clippy} checks on Ubuntu.
    \item Full test suite (\texttt{cargo test --all-targets}) on both Ubuntu
          and macOS.
    \item Documentation build (\texttt{cargo doc --no-deps}) with
          \texttt{\#![warn(missing\_docs)]} enforcement.
    \item All example programs are compiled and run on Ubuntu.
\end{itemize}

This provides cross-platform validation for the unsafe FFI and
platform-gated features (epoll/kqueue/poll, CPU affinity, io\_uring).


\section{Documentation conventions}

Every source file carries a \texttt{//!} module-level comment describing its
purpose and the types it contains. The crate root declares
\texttt{\#![warn(missing\_docs)]}, which makes undocumented public items
compilation errors.

\begin{itemize}
    \item \textbf{Public API} (\texttt{///}): required on all \texttt{pub}
          items. Enforced by the lint.
    \item \textbf{Module docs} (\texttt{//!}): present on every \texttt{.rs}
          file, both top-level public modules and internal sub-files.
    \item \textbf{Unsafe blocks} (\texttt{// SAFETY:}): every unsafe block
          has a preceding comment explaining why the operation is sound.
    \item \textbf{FFI blocks}: all use \texttt{unsafe extern "C" \{ ... \}}
          per Rust Edition 2024.
    \item \textbf{Test names}: follow a descriptive
          \texttt{subject\_action\_expected\_outcome} convention.
\end{itemize}

The standard commands for checking these are part of the normal validation
workflow:

\begin{lstlisting}[language=bash]
cargo fmt --check
cargo clippy --all-targets -- -D warnings
cargo test
cargo doc --no-deps
\end{lstlisting}

\section{Validation strategy}

Use narrow tests first, then broaden. This keeps iteration fast while still
protecting the runtime invariants.

\begin{lstlisting}[language=bash,caption={Typical focused-to-broad workflow}]
cargo test task_snapshot -- --nocapture
cargo test scheduling_group -- --nocapture
cargo test serve_tcp -- --nocapture
cargo fmt --check
cargo test
cargo clippy --all-targets -- -D warnings
cargo doc --no-deps
\end{lstlisting}

For changes touching Linux-only code, use the Docker helper after local macOS
validation:

\begin{lstlisting}[language=bash,caption={Linux platform pass}]
SITAS_DOCKER_IO_URING=1 tools/linux-docker.sh cargo test
SITAS_DOCKER_IO_URING=1 tools/linux-docker.sh cargo clippy --all-targets -- -D warnings
\end{lstlisting}

The order matters. Focused tests make it easier to understand the mechanism
that failed. Broad tests then protect against regressions in adjacent runtime
paths. This is especially important in executor code, where a small lifecycle
change can affect timers, readiness, joins, and shutdown at the same time.

\section{Interpreting unsupported features}

Unsupported is a valid result when the platform cannot provide a feature. CPU
placement is unsupported on non-Linux platforms. \texttt{io\_uring} may be
unavailable if the host kernel, container runtime, or security profile blocks
ring setup. The project should report those cases honestly instead of hiding
them behind a fake success path.

This matters for learning. A runtime experiment should make platform boundaries
visible: readiness works on several Unix backends, hard CPU affinity is
Linux-specific, and \texttt{io\_uring} is both Linux-specific and dependent on
host/container permissions.
