\chapter{Guided Walkthroughs}
\label{chap:walkthroughs}

\section{Purpose}

This chapter gives concrete source-reading paths. Each walkthrough starts from
a visible API or example and follows the mechanism through the runtime.

\begin{notabene}[How to use these walkthroughs]
Keep the source tree open while reading. Read the public function first, then
follow the listed modules in order.
\end{notabene}

\section{Walkthrough: spawning one task}

Start with the smallest async unit: a task spawned onto one executor.

\begin{lstlisting}[language={},caption={Task spawn reading path}]
examples/executor_sleep.rs
src/executor.rs
src/executor/spawner.rs
src/executor/task.rs
src/executor/task_state.rs
src/executor/scheduler.rs
src/executor/task_set.rs
\end{lstlisting}

The public API is \texttt{executor\_and\_spawner}. It returns one executor
and one spawner. The spawner creates tasks; the executor polls them.

The useful questions to ask while reading are:

\begin{enumerate}
    \item Where is the future boxed and pinned?
    \item Where does the task get its id and optional name?
    \item What prevents repeated wakes from putting the same task into the
          ready queue many times?
    \item What state is recorded before and after each poll?
    \item What happens if the executor is dropped before the task completes?
\end{enumerate}

The mechanism is local. A task spawned by a \texttt{Spawner} is owned by that
spawner's executor. Its waker requeues it there. Nothing in this path migrates
the task to another thread.

\section{Walkthrough: sleeping and timers}

The timer path starts with examples such as \filepath{executor_sleep.rs} and
\filepath{executor_timeout.rs}. Then read:

\begin{lstlisting}[language={},caption={Timer reading path}]
src/executor/future.rs
src/executor/timer.rs
src/executor/scheduler.rs
src/executor/driver.rs
\end{lstlisting}

\texttt{sleep} creates a future that registers a deadline. When polled before
the deadline, the future stores the task waker and returns
\texttt{Poll::Pending}. The scheduler keeps timer state separate from the ready
queues. On each loop iteration the executor checks expired timers and wakes the
registered tasks.

\texttt{timeout} demonstrates cancellation by drop. If the deadline wins, the
inner future is dropped. Any timer or readiness registration owned by that
future must clean itself up.

\section{Walkthrough: readiness-driven TCP}

Start from the server helpers rather than the raw reactor:

\begin{lstlisting}[language={},caption={TCP readiness reading path}]
examples/async_tcp_server.rs
examples/async_tcp_scoped_server.rs
src/executor/tcp.rs
src/executor/unix_io.rs
src/executor/io_interest.rs
src/os/reactor.rs
\end{lstlisting}

The TCP helper shows the application shape: accept, spawn a handler, wait for
handlers, propagate errors. The lower layers show the runtime shape: try a
non-blocking operation, register interest on \texttt{WouldBlock}, sleep in the
OS reactor, wake the task on readiness.

When reading \filepath{src/executor/tcp.rs}, notice that the accept loop and
handler tasks are distinct. Grouped TCP helpers use this: the accept loop stays
in the caller's task while accepted handler tasks are spawned into a scheduling
group.

\section{Walkthrough: scheduling groups}

The fastest way to see scheduling groups:

\begin{lstlisting}[language=bash,caption={Scheduling group examples}]
cargo run --example scheduling_group_demo -- 1
cargo run --example async_tcp_scheduling_groups
cargo run --example sharded_scheduling_groups
\end{lstlisting}

Then read:

\begin{lstlisting}[language={},caption={Scheduling group reading path}]
src/executor/scheduling_group.rs
src/executor/spawner.rs
src/executor/task_set.rs
src/executor/snapshot.rs
src/sharded_executor.rs
\end{lstlisting}

\texttt{scheduling\_group.rs} defines the public handle and its executor
ownership. \texttt{spawner.rs} validates that a group belongs to the executor
before creating a task in that group. \texttt{task\_set.rs} owns the queues and
weighted virtual runtime accounting. \texttt{snapshot.rs} resolves group names
into task snapshots for observability.

On the sharded runtime, the extra step is runtime identity. A
\texttt{ShardedSchedulingGroup} is a bundle of executor-local groups, one per
shard. It must not be accepted by a different runtime even if names and ids
look similar.

\section{Walkthrough: cross-shard submission}

Start with:

\begin{lstlisting}[language={},caption={Cross-shard reading path}]
examples/sharded_submit.rs
examples/sharded_map_reduce.rs
src/sharded_executor.rs
src/sharded_executor/join.rs
src/executor/spawner.rs
\end{lstlisting}

A cross-shard submit is not a function call into borrowed remote state. It is
owned work sent to another executor shard. The submitted future is polled on
the target shard. Its owned output completes a join handle. The original task
wakes on its original shard.

While reading the submitter APIs, look for the \texttt{Send + 'static}
boundaries that mark where work can cross threads.

\section{Walkthrough: shard-local state}

Read \filepath{examples/shard_local.rs} first, then:

\begin{lstlisting}[language={},caption={Shard-local reading path}]
src/shard_local.rs
src/sharded_executor.rs
src/runtime.rs
\end{lstlisting}

\texttt{ShardLocal<T>} is the clearest example of the shared-nothing idea in
the async runtime. Each shard owns one \texttt{T}. Access is routed to the
owning shard. The closure receives \texttt{\&mut T} and returns an owned result.

The key rule is negative: a reference to \texttt{T} must not become part of a
future that later awaits. The API enforces this through synchronous closures
rather than async callbacks.

\section{Walkthrough: \texttt{io\_uring} lifecycle}

This path is Linux-specific. When available, start with:

\begin{lstlisting}[language=bash,caption={io\_uring examples}]
SITAS_DOCKER_IO_URING=1 tools/linux-docker.sh cargo run --example os_uring
SITAS_DOCKER_IO_URING=1 tools/linux-docker.sh cargo run --example os_uring_abandon
\end{lstlisting}

Then read:

\begin{lstlisting}[language={},caption={io\_uring reading path}]
src/os/uring.rs
src/executor/uring.rs
src/executor/driver.rs
src/executor/tests.rs
\end{lstlisting}

The key distinction is ownership. A readiness future borrows a descriptor and
retries ordinary system calls. An \texttt{io\_uring} future submits owned
buffers to the kernel and must keep them alive until completion or safe
abandonment.

\filepath{os_uring_abandon.rs} is useful because success paths can hide
lifetime complexity. Abandonment forces the dispatcher to keep enough state to
clean up operations whose futures no longer exist.

\section{Walkthrough: reading observability output}

Run:

\begin{lstlisting}[language=bash,caption={Observability examples}]
cargo run --example sharded_observability
cargo run --example shard_local_worker_observability
\end{lstlisting}

The columns are a compact runtime story:

\begin{description}
    \item[\texttt{ready}] how many tasks are queued to be polled;
    \item[\texttt{tasks}] how many tasks are still unfinished;
    \item[\texttt{timers}] how many timer registrations exist;
    \item[\texttt{polls}] how many task polls the executor has performed;
    \item[\texttt{done}] how many tasks completed or were cancelled;
    \item[\texttt{age\_ms}] how long the task has existed;
    \item[\texttt{state\_ms}] how long it has been in the current state;
    \item[\texttt{wait}] the last known wait interest.
\end{description}

The output is not tracing. There is no event stream. It is repeated sampling of
owned snapshots---less powerful than a full console, but simple, portable, and
keeps runtime internals inspectable.

\section{Walkthrough: sharded index build}

The index build is the most application-like example. Start with a small run:

\begin{lstlisting}[language=bash,caption={Index build example}]
cargo run --example sharded_index_build -- --records 512 --shards 4 --seed 0xbeef --cleanup
\end{lstlisting}

Then read:

\begin{lstlisting}[language={},caption={Index build reading path}]
examples/sharded_index_build.rs
src/shard_local.rs
src/sharded_executor.rs
src/running_stats.rs
\end{lstlisting}

The example demonstrates several runtime mechanisms working together:

\begin{enumerate}
    \item \texttt{ShardLocal<ShardProgress>} holds per-shard progress
          counters. Each shard task calls \texttt{with\_current\_result}
          to update only its own slot without a shared mutex.
    \item \texttt{runtime.map\_all} dispatches partition scanning to all
          shards in one call, replacing the older pattern of collecting
          join handles one at a time.
    \item Chunked reads (256 records per \texttt{read\_exact}) replace
          one-syscall-per-record scanning.
    \item After all partitions complete, \texttt{RunningStatistics}
          collects per-shard scan durations and prints mean, standard
          deviation, CV, min, and max.
    \item The merge phase submits pairs of sorted run files back onto
          shards. Each merge task reads two sorted run files, merges them,
          and writes a new run file. The round continues until one sorted
          index remains.
    \item Verification reads every index entry and confirms global sort
          order and correct record offsets, but does so outside the
          sharded runtime.
\end{enumerate}

The runtime question is \emph{where} work runs. Partition tasks run on assigned shards. Merge tasks are submitted to specific shards
(the left input run's shard). Progress reads happen through
\texttt{ShardLocal::map\_all}, which submits synchronous closures to every
shard and collects owned results. At no point does a shard borrow another
shard's progress counter.

The \texttt{io\_uring} variant (\filepath{examples/sharded\_index\_build\_uring.rs})
replaces partition reads, run writes, and merge run I/O with
\texttt{read\_exact\_at\_uring} and \texttt{write\_all\_at\_uring} futures.
It keeps the same \texttt{ShardLocal}, \texttt{map\_all}, and
\texttt{RunningStatistics} shape.

\subsection{Mailbox-transfer index path}

Then run the mailbox variant:

\begin{lstlisting}[language=bash,caption={Mailbox index example}]
cargo run --example sharded_index_mailbox -- --records 512 --shards 4 --seed 0xbeef
\end{lstlisting}

Read:

\begin{lstlisting}[language={},caption={Mailbox index reading path}]
examples/sharded_index_mailbox.rs
src/shard_mailbox.rs
src/placement.rs
src/sharded_executor.rs
\end{lstlisting}

This variant keeps scan work sharded, but replaces intermediate file-mediated
exchange with typed mailbox messages. The key question changes from ``which
shard owns this key?'' to ``which target type does this router produce?''

\begin{itemize}
    \item \texttt{UniformShardRouter} produces a physical \texttt{ShardId}.
          That is the uniform route into \texttt{ShardMailboxSet<M>}.
    \item \texttt{WorkUnitRouter<IndexWorkUnit>} produces a logical assembler
          name. That is the non-uniform route into
          \texttt{WorkUnitMailboxSet<N, M>}.
    \item \texttt{WorkUnitSpec<N>} assigns each logical assembler to an
          executor shard. The default example assigns two assemblers to shard
          0, making non-uniform placement visible.
\end{itemize}

Keep the roles separate while reading the example. Scanner tasks message
assembler work units through mailboxes. Assemblers report completion metadata
to \texttt{main} through task handles, not through another mailbox, and their
sorted data reaches coordination as run files consumed by the final k-way
merge.

While reading \filepath{src/shard\_mailbox.rs}, distinguish the transport from
the ownership model. The mailbox queue uses synchronization internally, but
the message is an owned value and the receiver task is still the only code that
mutates the work unit's local assembly state.

\section{Debugging checklist}
\label{sec:debugging-checklist}

When something behaves unexpectedly, ask these questions in order:

\begin{enumerate}
    \item Which shard owns the state being mutated?
    \item Which executor polls the future?
    \item Is the task queued, polling, waiting, completed, or cancelled?
    \item If waiting, is it waiting for a timer, fd readiness, completion, or
          an unknown waker?
    \item Did a handle from another executor or runtime cross a boundary?
    \item If using \texttt{io\_uring}, who owns the buffer until completion?
    \item If shutdown hangs, which spawner, submitter, join handle, or scoped
          child is still alive?
\end{enumerate}

These questions match the runtime's invariants. They are usually more
productive than starting with performance guesses.
